Reproducible Research: Peer Assessment 1 - Activity Monitoring Data
===================================================================


## Loading the data set and the relevant libraries

```{r}
library(reshape2)
library(plyr)
library(lattice)

if(!file.exists('./activity.csv')){
     if(!file.exists('./activity.zip')){
          url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
          download.file(url, 'activity.zip', method=ifelse(.Platform$OS.type=="windows", "internal", "curl"))
     }
     unzip('activity.zip')
}
data <- read.csv('activity.csv')

```

## Reformatting the data set
```{r}
melted <- melt(data, id.vars=c("date"))
```

## Total number of steps taken each day
```{r steps_per_day, fig.width=14, fig.height=8}
casted <- dcast(melted[melted$variable=='steps',], date ~ variable, sum)
par(las=2, mgp=c(4,1,0), mar=c(8,8,4,1), cex.lab=1.2)
barplot(casted$steps, names.arg=casted$date, cex.names=.7)
```

### The mean of the total steps taken per day is *`r mean(casted$steps, na.rm=TRUE)`* and the median is *`r median(casted$steps, na.rm=TRUE)`*

intervalMelt <- melt(data, id.vars=c("interval"))


## Average daily activity pattern
```{r average_daily_activity, fig.width=14}
intervalMelt <- melt(data, id.vars=c("interval"))
intervalMelt <- intervalMelt[intervalMelt$variable == 'steps',]
class(intervalMelt$value) <- "integer"
intervalCasted <- dcast(intervalMelt, interval ~ variable, mean, na.rm=TRUE)

plot(intervalCasted$steps, type='l', xlab='Interval', ylab='Number of steps mean', axes=FALSE)

axis(1, at=1:288, labels=intervalCasted$interval)
```

## The 5-minute interval that contains the maximum number of steps (averaged accross days) is: *`r arrange(intervalCasted, desc(steps))[1,1]`*  

## Inputing missing values
The missing values will be replaced with the average of the corresponding 5-minutes interval (averaged accross days)

```{r}
completeCases = complete.cases(data)
cleanData <- data

for(i in 1:nrow(cleanData)){
     row = cleanData[i,]
     if(is.na(row$steps)){
          cleanData[i, 'steps'] = intervalCasted[intervalCasted$interval == row$interval, 'steps']
     }
}

```


## Plotting a histogram of the total number of steps per day after inputing the missing values
```{r steps_per_day_after_data_input, fig.width=14, fig.height=8}
melted <- melt(cleanData, id.vars=c("date"))
casted <- dcast(melted[melted$variable=='steps',], date ~ variable, sum)
par(las=2, mgp=c(4,1,0), mar=c(8,8,4,1), cex.lab=1.2)
barplot(casted$steps, names.arg=casted$date, cex.names=.7)
```

### The mean of the total steps taken per day is *`r mean(casted$steps, na.rm=TRUE)`* and the median is *`r median(casted$steps, na.rm=TRUE)`*
The values are the same as the values obtained with ra.rm = TRUE  
Which is mathematically correct since the values we added are equal to the mean of each corresponding 5-minute interval 


## Difference in activity patterns between weekdays and weekends
```{r  weekend_vs_weekday, fig.width=13, fig.height=16}
cleanData <- data
melted <- melt(cleanData, id.vars=c("date", "interval"))
melted$date = as.Date(melted$date)
melted <- cbind(melted, weekdays(melted$date))
colnames(melted)[5] <- 'weekend'
melted$weekend <- melted[,'weekend'] == 'Saturday' | melted[,'weekend'] == 'Sunday' 

casted <- dcast(melted[melted$variable=='steps',], interval + weekend ~ variable, mean, na.rm=TRUE)
casted[casted$weekend, 'weekend'] = 'weekend'
casted[casted$weekend == 'FALSE', 'weekend'] = 'weekday'

xyplot(steps ~ interval | weekend, data=casted, type='l', xlab='Interval', ylab='Number of steps')

```

